# Distributions {#sec-summaries-distributions}

```{r}
#| cache: false
#| warning: false
#| message: false
#| echo: false

library(tidyverse)
library(behdslabs)
library(scales)
```

Before we begin, we clarify that this chapter will not cover distributions in the context of probability, which is the mathematical framework used to model uncertainty. We postpone that discussion until the next part of the book, where we will use it throughout. Here, we focus on _distributions_ as a way to describe and summarize data we've already collected. Our goal is to understand the variation and patterns present in a dataset, without yet making probabilistic claims or predictions. The concepts introduced in this chapter will later serve as the foundation for understanding probability distributions, where similar tools are used to describe the likely outcomes of random processes.

To illustrate the concepts needed to understand distributions and how they relate to summary statistics, we will pretend that we have to describe the reaction times of our study participants to ET, an extraterrestrial that has never run a psychology experiment. As a first step, we need to collect data. To do this, we ask participants to complete a simple reaction-time task and record their response time in milliseconds. We ask them to provide sex information because we want to compare the two groups. We collect the data and save it in the `reaction_times` data frame included in the **behdslabs** package.

```{r}
library(behdslabs)
str(reaction_times)
```


One way to convey the reaction times to ET is to simply send the entire list of measurements. However, there are much more effective ways to convey this information, and understanding the concept of a distribution will help. To simplify the explanation, we first focus on male reaction times. We examine the female reaction-time data in @sec-stratification.

It turns out that, in some cases, the two summary statistics introduced in the next chapter, the average and the standard deviation, are all we need to understand the data. We will learn data visualization techniques that will help us determine when this two-number summary is appropriate. These same techniques will serve as an alternative for when two numbers are not enough.

## Variable types 

Knowing what variable type we are working with is important because it will help us determine the most effective way to summarize our data and display its distribution.

We will be working with two types of variables: categorical and numeric. Each can be divided into two other groups: categorical can be ordinal or not, whereas numerical variables can be discrete or continuous.

When data values represent groups instead of numbers, we refer to the data as _categorical_. Two simple examples are sex (male or female) and US regions (Northeast, South, North Central, West). Some categorical data can be ordered even if they are not numbers, such as spiciness (mild, medium, hot). These ordered categorical data are referred to as _ordinal_ data.

Numerical data can be continuous or discrete. A continuous variable can take on any value within a range. For example, if measured with sufficient precision, reaction time is continuous because the difference between two participants (even identical twins) can be an arbitrarily small fraction of a millisecond. However, if reaction time is rounded to the nearest millisecond, it becomes discrete since values must be whole numbers. Other examples of discrete data are the number of heads when tossing 10 coins, the number of cigarettes a person smokes per day, or the number of customers visiting a store in an hour. In these cases, the values are countable and cannot take on fractional values. Unlike continuous data, which can be measured with increasing precision, discrete data are limited to distinct, separate values.

In practice, there are occasions in which discrete variables are treated as continuous. For example, a digital measuring instrument might record data as 16-bit integers. However, because it is rare to obtain exactly the same reading more than once, it is convenient to treat the data as continuous. Another example is country population size. Although population counts are always whole numbers, it is uncommon for two jurisdictions to have exactly the same population. For this reason, population size is also commonly treated as continuous.


## Relative frequency distribution {#sec-pmf-intro}

The most basic statistical summary of a list of numbers or categories is its *distribution*. By a *list*, we simply mean a collection of observed values, such as a column in a dataset. The order of the values does not matter, but repeated values are allowed. The simplest way to think of a distribution is as a compact description of how values are spread across observations. This concept should not be new for readers of this book. For example, with categorical data, the _relative frequency distribution_ simply describes the proportion of each unique category:

$$
p_k = \frac{\mbox{number of times category }k \mbox{ appears in the list}}{n}
$$

with $n$ as the length of the list.

Here is an example with US state regions:

```{r summaries-state-region-distribution}
prop.table(table(state.region))
```

When the data are discrete numerical values, the relative frequency distribution is defined in the same way, except it is more common to use the following notation since the frequency can be considered a function of numerical values:

$$
f(x) = \frac{\mbox{number of times value } x \mbox{ appears in the list}}{n}
$$


Here is the distribution for US state populations rounded to the nearest million, which are discrete numerical values:

```{r summaries-rounded-population-distribution}
#| echo: false
app_data_breaches |> mutate(x = round(population/10^6)) |> count(x) |> mutate(f = n/sum(n)) |>
  ggplot(aes(x, f)) + geom_col() + 
  scale_x_continuous(breaks = seq(1, 37, by = 1)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5, size = 8)) +
  ylab("f(x)") + theme(panel.grid = element_blank())
```

## Empirical cumulative distribution functions {#sec-ecdf-intro}

When the data is continuous, the task of constructing a summary based on a distribution is more challenging because reporting the frequency of each entry is not an effective summary: most entries are unique so the relative frequency distribution does not summarize much.

Here are the relative frequencies of US state populations, no longer rounded:

```{r relative-frequencies-of-US}
#| echo: false
app_data_breaches |> filter(state %in% state.name) |>
  count(population) |> mutate(f = n/sum(n)) |> ggplot(aes(population, f)) + 
  geom_col() + xlab("x = population") + ylab("f(x)") + theme(panel.grid = element_blank()) +
  ylab("f(x)") + theme(panel.grid = element_blank()) + 
  scale_x_continuous(labels = label_comma())
```

Since there are 50 states, each with a unique population value $x$, the plot shows $f(x) = \frac{1}{50} = 0.02$ for each state.

To define a distribution for continuous numeric data, we define a function that reports the proportion of the data entries $x$ that are below a threshold $a$, for all possible values of $a$. This function is called the empirical cumulative distribution function (eCDF) and often denoted with $F$:

$$ F(a) = \mbox{Proportion of data points that are less than or equal to }a$$

Here is the eCDF for state population data above:

```{r state-population-ecdf}
#| echo: false
app_data_breaches |> filter(state %in% state.name) |>
  ggplot(aes(population)) + 
  stat_ecdf() +
  labs(x = "a", y = "F(a)") + 
  scale_x_continuous(labels = label_comma())
```

Similar to how a frequency table summarizes categorical data, the empirical cumulative distribution function (eCDF) summarizes continuous data. The eCDF provides a clear picture of how the values are distributed and highlights key characteristics of the dataset. For instance, in the population data, we can see that about half of the states have populations greater than 5,000,000, while most states fall below 20,000,000. This type of summary helps us to quickly identify medians, ranges, and other important distributional features.

In our reaction-time case study, response times were recorded with sub-millisecond precision, so almost every value is unique. For example, one participant recorded a reaction time of `283.1325` milliseconds and another recorded `435.0265` milliseconds, and no other participant matched either value exactly.

As a result the data is continuous and the relative frequency distribution is somewhat confusing:

```{r frequency-of-continuous-data-not-good}
#| echo: false
reaction_times |> filter(sex == "Male") |> 
  count(rt_ms) |> mutate(f = n/sum(n)) |> ggplot(aes(rt_ms, f)) + 
  geom_col(width = 0.1) +
  xlab("x = reaction time (ms)") + ylab("f(x)")
```

Although this approach preserves all the information from the list of observations, it can give misleading impressions if extrapolating insights to a broader population. For example, it might suggest that participants are much more likely to respond in exactly 374.4 milliseconds than in 374.0 milliseconds, or that no participant responds between 350 and 351 milliseconds, patterns we know aren't accurate. In contrast, the eCDF provides a more useful visual summary:

```{r summaries-ecdf}
#| echo: false
reaction_times |> filter(sex == "Male") |> 
  ggplot(aes(rt_ms)) + 
  stat_ecdf() +
  labs(x = "a", y = "F(a)")
```

From a quick glance at the figure, we can already infer some general characteristics of the data. For example, about half of the males respond in under 375 milliseconds, since $F(375) \approx 0.5$, and responding in under 275 milliseconds or taking more than 475 milliseconds is very rare. Moreover, because we can compute the proportion of participants with reaction times between any two values $a$ and $b$ as $F(b) - F(a)$, the plot above contains all the information needed to reconstruct the entire dataset.

Paraphrasing the saying “a picture is worth a thousand words,” in this case, a picture is as informative as `r sum(reaction_times$sex == "Male")` numbers.

:::{.callout-note}
The reason we add the word *empirical* is because, as we will see in @sec-cdf-intro, the cumulative distribution function (CDF) can be defined mathematically, meaning without any data.
:::

## Histograms

Although the eCDF is a fundamental concept in statistics, it is not commonly used in practice. The main reason is that it does not easily highlight key distribution characteristics, such as the central tendency, symmetry, or the range that contains 95% of the values. In contrast, histograms are widely preferred because they make these features much more interpretable. While histograms sacrifice a small amount of information, they provide a much clearer summary of the data.

To construct a histogram, we first divide the range of the data into non-overlapping bins. Each bin acts as a category, and its height is chosen so that the area of the bar, calculated as the height multiplied by the bin width, represents the relative frequency of observations within that bin. Although a histogram resembles a bar plot, it differs in that the x-axis represents numerical values rather than categorical ones.

If we round the reaction times to the nearest 10 milliseconds and create a relative frequency distribution, we obtain a histogram with bins defined in 10-millisecond intervals: 

$$(195, 205], (205, 215], (215, 225], (225, 235], \dots, (495, 505]$$

We can do this with the following **ggplot2** code:

```{r frequency-distribution-rounded-rt}
library(tidyverse)
reaction_times |> filter(sex == "Male") |> 
  mutate(rt_ms = round(rt_ms, -1)) |>
  count(rt_ms) |> mutate(f = n / sum(n)) |> 
  ggplot(aes(rt_ms, f)) + geom_col(width = 8)
```

The R base `hist` function also allows us to quickly construct a histogram. We can define custom bin intervals which is particularly useful for grouping extreme values and avoiding empty bins:

```{r hist-custom-bin-example}
with(reaction_times, hist(rt_ms[sex == "Male"], breaks = c(200, 250, seq(300, 450, by = 10), 500)))
```

If we were to send either of these plots to an extraterrestrial unfamiliar with human reaction times, they would immediately gain key insights about the data. First, the male reaction times range from 200 to 500 milliseconds, with over 95% of observations falling between 305 and 439 milliseconds. Second, the distribution is approximately symmetric around 375 milliseconds. Additionally, by summing bin counts, one could estimate the proportion of data falling within any given interval. Thus, a histogram provides an intuitive summary that preserves most of the information from the original `r sum(reaction_times$sex=="Male")` reaction-time measurements using only about 30 bin counts.

What information is lost?  A histogram treats all values within a bin as if they were the same when computing bar heights. For instance, values of 340.0, 340.1, and 340.2 milliseconds are grouped together. However, since these differences are barely perceptible, the practical impact is negligible, and we achieve a meaningful summary using just 18 numbers.

:::{.callout-note}
The `reaction_times` data are simulated, and their distribution is close to symmetric. Real reaction times are typically **right-skewed**: most responses are fairly quick, but a minority are much slower (lapses of attention, for example), producing a long right tail. Keep this in mind when we describe distribution shapes below, and do not take the symmetry of this dataset as a general property of reaction-time data.
:::

By default, histograms often display frequency on the y-axis instead of relative frequency. The **ggplot2** package follows this convention:

```{r summaries-rt-histogram}
reaction_times |> filter(sex == "Male") |> ggplot(aes(x = rt_ms)) + 
  geom_histogram(breaks = c(200, 250, seq(300, 450, by = 10), 500), color = "black")
```

Note we use `fill` and `color` to mimic the behavior of `hist`.

When displaying frequencies, it is important to use equally spaced bins, as larger bins will naturally contain more observations and may distort the visual representation of the data. To instead display **relative** frequency, as done by `hist` above, you can modify the code by setting `y = after_stat(density)`:

```{r summaries-rt-histogram-ggplot2}
#| eval: false
reaction_times |> filter(sex == "Male") |> 
  ggplot(aes(x = rt_ms, y = after_stat(density))) + 
  geom_histogram(breaks = c(200, 250, seq(300, 450, by = 10), 500), color = "black")
```


### Choosing bin widths

The appearance and interpretation of a histogram depend on the choice of bins, and in the case of equally sized bins, their width. Different bin widths can make the same data look quite different: too narrow, and the histogram may appear noisy or fragmented; too wide, and important features of the distribution may be hidden. The choice of bin width should be guided by the context of the data. For example, in the case of reaction times, timing hardware and software have limited resolution, so using bins much smaller than about 10 milliseconds does not make sense. At the same time, because differences of more than 50 milliseconds are meaningful in a psychological experiment, bins larger than about 20 milliseconds would obscure meaningful variation. For the extreme regions of the distribution, such as below 300 milliseconds or above 450 milliseconds, where there are few observations, it is reasonable to use wider bins to keep the summary clear. Every dataset requires a context-specific decision: the goal is to choose bins that reveal structure without introducing noise, maximizing the information conveyed by a single figure.

## Smoothed density plots

Smooth density plots are similar to histograms, but without bins. Instead of sharp edges, they produce a continuous curve:

```{r summaries-example-of-smoothed-density}
#| echo: false
reaction_times |> filter(sex == "Male") |> 
  ggplot(aes(rt_ms)) + geom_density(alpha = 0.2, fill = "#00BFC4")
```

To understand smooth densities, think of our sample as coming from a much larger, unobserved population. If we had many more observations, we could build a histogram with very small bins. As bin width decreases, the histogram becomes smoother and approaches a continuous curve:

```{r summaries-simulated-data-histogram-2, fig.width = 9, fig.height = 3,  out.width = "100%",echo = FALSE, message = FALSE}
set.seed(1988)
x <- data.frame(rt_ms = c(rnorm(1000000, 377, 33), rnorm(1000000, 337, 33)))
p1 <- x |> 
  ggplot(aes(x = rt_ms, y = after_stat(density))) + 
  geom_histogram(binwidth = 10, fill = "grey", color = "black") + 
  ggtitle("binwidth = 10")
p2 <- x |> 
  ggplot(aes(x = rt_ms, y = after_stat(density))) + 
  geom_histogram(binwidth = 5, fill = "grey", color = "black") + 
  ggtitle("binwidth = 5")
p3 <- x |> 
  ggplot(aes(x = rt_ms, y = after_stat(density))) + 
  geom_histogram(binwidth = 2.5, fill = "grey", color = "black") + 
  ggtitle("binwidth = 2.5")
library(gridExtra)
grid.arrange(p1, p2, p3, nrow = 1)
```

A smooth density can be viewed as the curve tracing the tops of these increasingly fine histogram bars.

In practice, we have limited data, so we use a histogram with reasonable bin width and overlay a smooth curve. The following steps illustrate how this curve is constructed:

```{r summaries-smooth-density-2, echo = FALSE, out.width = "100%"}
hist1 <- reaction_times |> 
  filter(sex == "Male") |> 
  ggplot(aes(rt_ms)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 10, fill = "grey", color = "black") 

hist2 <- hist1 +
  geom_line(stat = 'density', lwd = 0.75)

hist3 <- hist1 + 
  geom_point(data = ggplot_build(hist2)$data[[1]], aes(x,y), col = "blue")

hist4 <- ggplot() + geom_point(data = ggplot_build(hist2)$data[[1]], aes(x,y), col = "blue") + 
  xlab("reaction time (ms)") + ylab("density")

hist5 <- hist4 + geom_line(data = ggplot_build(hist2)$data[[2]], aes(x,y), lwd = 0.75)

hist6 <- reaction_times |> 
  filter(sex == "Male") |> 
  ggplot(aes(rt_ms)) +
  geom_density(alpha = 0.2, fill = "#00BFC4", col = 0) +
  geom_line(stat = 'density', lwd = 0.75) +
  scale_y_continuous(limits = layer_scales(hist2)$y$range$range)
  
grid.arrange(hist1, hist3, hist4, hist5, hist2, hist6, nrow = 2)
```

### Choosing the bandwidth

The smoothness of a density curve is controlled by a parameter called the *bandwidth*. Smaller values produce more detail (and noise), while larger values produce smoother curves:

```{r summaries-densities-different-smoothness, out.width = "100%", fig.width = 6, fig.height = 3, echo=FALSE,cache=FALSE, message=FALSE}
p <- reaction_times |> filter(sex == "Male") |> 
  ggplot(aes(rt_ms)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 10, alpha = 0.5) 

p1 <- p +  geom_line(stat = 'density', adjust = 0.5)
p2 <- p +  geom_line(stat = 'density', adjust = 2) 

library(gridExtra)
grid.arrange(p1, p2, ncol = 2)
```

As with bin width in histograms, the goal is to balance noise and smoothness.

### Interpreting the y-axis

The y-axis is scaled so the total area under the curve equals 1. Probabilities are represented by areas under the curve. For example, the proportion of values between 350 and 380 milliseconds is the area over that interval:

```{r summaries-area-under-curve, echo = FALSE}
d <- with(reaction_times, density(rt_ms[sex == "Male"]))
tmp <- data.frame(rt_ms = d$x, density = d$y)
tmp |> ggplot(aes(rt_ms,density)) + geom_line() + 
  geom_area(aes(x = rt_ms, y = density), data = filter(tmp, between(rt_ms, 350, 380)), alpha = 0.2, fill = "#00BFC4")
```

This proportion is about `r round(mean(dplyr::between(reaction_times$rt_ms[reaction_times$sex=="Male"], 350, 380)), 2)`, meaning roughly `r noquote(paste0(round(mean(dplyr::between(reaction_times$rt_ms[reaction_times$sex=="Male"], 350, 380)), 2)*100, '%'))` of male reaction times fall in this range.

With this interpretation, smooth densities provide a useful and visually appealing summary of the data:

```{r summaries-example-of-smoothed-density-2, echo = FALSE}
reaction_times |> 
  filter(sex == "Male") |> 
  ggplot(aes(rt_ms)) + 
  geom_density(alpha = 0.2, fill = "#00BFC4")
```

:::{.callout-note}
With the material covered up to this point, you can complete exercises 1 through 10.
:::

## Describing distributions

When visualizing data with a histogram or density plot, we often describe the shape of a distribution using a few key terms and concepts. In this section, our goal is simply to familiarize you with this vocabulary, terms that you'll encounter in most discussions of data analysis.

A _mode_ is a value or region where the data are most concentrated, the peak of the distribution. Some distributions have one peak (unimodal), while others have two or more (bimodal or multimodal), which can indicate distinct subgroups within the data.

A _symmetric_ distribution has as many data points to both sides of the mode. A distribution is _skewed_ if it is asymmetric.

The _tails_ of a distribution are the regions that extend toward the smallest and largest values.

A _right-skewed_ (positively skewed) distribution has a long tail toward larger values, for example, income, where most people earn modest amounts but a few earn much more. Real reaction times are another example: most responses are quick but a few are very slow.

A _left-skewed_ (negatively skewed) distribution has a longer tail toward smaller values, such as test scores on an easy exam where most students score high but a few perform poorly.

In the next chapter, we introduce the normal distribution, a symmetric, _bell-shaped_ distribution that serves as a useful reference point. Distributions with _heavy tails_ contain more extreme observations than the normal distribution, while light-tailed distributions contain fewer.

Here are some simulated examples illustrating the terms described above:

```{r distribution-shapes, echo=FALSE, fig.height=5, out.width="100%"}
library(ggplot2)
library(dplyr)
set.seed(42)

df <- data.frame(
  value = c(rnorm(10000, 0, 0.75),
            rexp(10000, 1) - 2.5,
            -rexp(10000, 1) + 2.5,
            c(rnorm(9000, -1, 0.5), rnorm(1000, 2, 0.5))),
  type = rep(c("Bell shaped", "Right-skewed", "Left-skewed", "Bimodal"), each = 10000)) |>
  mutate(type = factor(type, levels = c("Bell shaped","Bimodal","Right-skewed","Left-skewed")))


ggplot(df, aes(value, fill = type)) +
  geom_density(alpha = 0.6, color = "black", adjust = 2, show.legend = FALSE) +
  facet_wrap(~type, ncol = 2, scales = "free_y") +
  theme_minimal() +
  coord_cartesian(xlim = c(-3.5, 3.5)) +
   theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(y = "", x = "")
```

## From visual to numerical summaries

As we've seen, visual tools such as histograms and density plots help us understand the shape of a distribution, its center, spread, symmetry, and tails. However, graphical summaries are often complemented by numerical summaries, which distill much of this information into just a few key numbers. In the next chapter, we introduce two of the most useful numerical summaries, the average and the standard deviation, which provide all the information you need when the data are approximately normal. We will also explore alternative measures that are more appropriate when distributions deviate from normality. The visualization techniques introduced here will guide us in deciding when simple numerical summaries are sufficient and when more detailed descriptions are needed.

## Exercises

1\. In the `app_data_breaches` dataset, the region is a categorical variable and the following is its distribution:

```{r summaries-barplot-exercise, echo = FALSE}
library(behdslabs)
ds_theme_set()
app_data_breaches |> group_by(region) |>
  summarize(n = n()) |>
  mutate(Proportion = n/sum(n), 
         region = reorder(region, Proportion)) |>
  ggplot(aes(x = region, y = Proportion, fill = region)) + 
  geom_bar(stat = "identity", show.legend = FALSE) + 
  xlab("")
```

To the closest 5%, what proportion of the states are in the North Central region?

2\. Which of the following is true:

a.  The graph above is a histogram.
b.  The graph above shows only four numbers with a bar plot.
c.  Categories are not numbers, so it does not make sense to graph the distribution.
d.  The colors, not the height of the bars, describe the distribution.

3\. The plot below shows the eCDF for male reaction times:

```{r summaries-ecdf-exercise, echo = FALSE}
reaction_times |> filter(sex == "Male") |> ggplot(aes(rt_ms)) + 
  stat_ecdf() +
  ylab("F(a)") + xlab("a")
```

Based on the plot, what percentage of males have a reaction time under 425 milliseconds?

a.  100%
b.  95%
c.  80%
d.  425 milliseconds

4\. To the closest 25 milliseconds, what reaction time $m$ has the property that 1/2 of the male participants are slower than $m$ and 1/2 are faster?

a.  250 milliseconds
b.  300 milliseconds
c.  375 milliseconds
d.  450 milliseconds

5\. Here is an eCDF of the incident rates (number of reported data-breach incidents per 100,000 people) across states:

```{r summaries-ecdf-exercise-2, echo = FALSE}
app_data_breaches |> mutate(incident_rate = incidents/population * 10^5) |>
  ggplot(aes(incident_rate)) + 
  stat_ecdf() +
  ylab("F(a)") + xlab("a")
```

Knowing that there are 51 observations (50 states and DC) and based on this plot, how many states have incident rates larger than 20 per 100,000 people?

a.  1
b.  5
c.  10
d.  50

6\. Based on the eCDF above, which of the following statements are true:

a.  About half the states have incident rates above 14 per 100,000 and the other half below.
b.  Most states have incident rates below 4 per 100,000.
c.  All the states have incident rates above 4 per 100,000.
d.  With the exception of 4 states, the incident rates are below 10 per 100,000.

7\. Below is a histogram of male reaction times in our `reaction_times` dataset:

```{r summaries-rt-histogram-exercise, echo = FALSE}
reaction_times |> 
  filter(sex == "Male") |> 
  ggplot(aes(rt_ms)) + 
  geom_histogram(binwidth = 10, color = "black")
```

Based on this plot, how many males have reaction times between 345 and 365 milliseconds?

a.  15
b.  50
c.  150
d.  500

8\. About what **percentage** have reaction times under 300 milliseconds?

a.  2%
b.  10%
c.  25%
d.  50%

9\. Based on the density plot below, about what proportion of US states have populations larger than 10 million?

```{r summaries-density-exercise, echo = FALSE}
app_data_breaches |> ggplot(aes(x = population/10^6)) + 
  geom_density(fill = "grey") + 
  scale_x_log10() +
  xlab("Population in millions")
```

a.  0.02
b.  0.15
c.  0.50
d.  0.55

10\. Below are three density plots. Is it possible that they are from the same dataset?

```{r summaries-density-exercise-2, echo = FALSE, warning = FALSE, message = FALSE}
library(gridExtra)
p1 <- app_data_breaches |> ggplot(aes(x = population/10^6)) + 
  geom_density(fill = "grey", bw = 5) + xlab("Population in millions") + ggtitle("1")
p2 <- app_data_breaches |> ggplot(aes(x = population/10^6)) + 
  geom_density(fill = "grey", bw = .05) + scale_x_log10() + xlab("Population in millions") + ggtitle("2")
p3 <- app_data_breaches |> ggplot(aes(x = population/10^6)) + 
  geom_density(fill = "grey", bw = 1) + scale_x_log10() + xlab("Population in millions") + ggtitle("3")
grid.arrange(p1,p2,p3,ncol = 2)
```

Which of the following statements is true:

a.  It is impossible that they are from the same dataset.
b.  They are from the same dataset, but the plots are different due to coding errors.
c.  They are the same dataset, but the first and second plot under smooth and the third over smooths.
d.  They are the same dataset, but the first is not in the log scale, the second under smooths, and the third over smooths.


